Which chemical property is related to white wine quality? by Zi Ye  
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
setwd('C:/Users/yz_333/Desktop/')
wine = read.csv('wineQualityWhites.csv', row.names = 1)
```

This is a dataset of white wine variants of the Portuguese "Vinho Verde" wine.

# Univariate Plots Section

A summary of this dataset including mean, median, and quantiles:

```{r echo=FALSE, Summary}
summary(wine)
```

For a better visualization of distributions, each variable is plotted in a 
histogram:

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram}
u1 = ggplot(aes(x = fixed.acidity), data = wine) + geom_histogram()
u2 = ggplot(aes(x = volatile.acidity), data = wine) + geom_histogram()
u3 = ggplot(aes(x = citric.acid), data = wine) + geom_histogram()
u4 = ggplot(aes(x = residual.sugar), data = wine) + geom_histogram()
u5 = ggplot(aes(x = chlorides), data = wine) + geom_histogram()
u6 = ggplot(aes(x = free.sulfur.dioxide), data = wine) + geom_histogram()
u7 = ggplot(aes(x = pH), data = wine) + geom_histogram()
u8 = ggplot(aes(x = total.sulfur.dioxide), data = wine) + geom_histogram()
u9 = ggplot(aes(x = density), data = wine) + geom_histogram()
u10 = ggplot(aes(x = sulphates), data = wine) + geom_histogram()
u11 = ggplot(aes(x = alcohol), data = wine) + geom_histogram()
u12 = ggplot(aes(x = quality), data = wine) + geom_histogram()
grid.arrange(u1, u2, u3,u4, u5, u6, u7, u8, u9, u10, u11, u12, ncol = 4)
```

# Univariate Analysis

### The structure:

This dataset contains 12 variables (including 11 chemical properties and the 
median quality evaluated by 3 experts) of 4898 white wine samples.

### The main feature that I am interested:

The most interesting feature is the quality of wines. I am interested in how 
chemical properties of white relate to their taste (quality). Acoording to the 
features of variables, acidity, sulfur dioxide, alcohol, and sugar content might
have impact on taste and quality.

### The distribution of variables:

All variables seem to have distributions that are close to normal. Residual 
sugar content has a slightly positive skew distribution. Sulphates content looks
like a bimodal distribution. Alcohol content seems to have a relatively large 
standard deviation.

### The adjustment I did:

I created a new variable called quality.factor by transforming the quality 
variable to factors since the quality score is discrete and each score is a 
evaluation category. I also noticed that the fixed acidity and volatile acidity
can be combined into a total.acidity variable to simplify the analysis.

```{r, New_variables}
wine = transform(wine, total.acidity = fixed.acidity + volatile.acidity)
wine = transform(wine, quality.factor = factor(quality))
```

# Bivariate Plots Section

Here I am interested to investigate which chemical property is related to the 
wine quality. According to the description of variables, high level of acidity 
and sulfur dioxide seem to give the wine bad taste, and it is not clear which 
properties are related to good taste.

A quick way to study the interaction between variables is to simply plot all of
them against each other and calculate their correlations:

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
ggpairs(wine, axisLabels="internal", 
        lower = list(continuous = wrap('points', shape = I('.'))),
        upper = list(continuous = wrap("cor", size = 3)))
```

Look like only alcohol content has a moderate correlation (0.436) with quality.
Here I plotted alcohol against quality.factor:

```{r echo=FALSE}
ggplot(aes(x = quality.factor, y = alcohol), data = wine) +
  geom_boxplot()
```

It is quite noticeable that high quality wine also has high alcohol content.
Then I plotted total.acidity and total.sulfur.dioxide that are suspected to
negatively impact the quality against the quality.factor:

```{r echo=FALSE}
ggplot(aes(x = quality.factor, y = total.acidity), data = wine) +
  geom_boxplot()
ggplot(aes(x = quality.factor, y = total.sulfur.dioxide), data = wine) +
  geom_boxplot()
```

It does not look like acidity and sulfur dioxide impact the quality. 
Then I plotted sugar which is a testant against quality.factor: 

```{r echo=FALSE}
ggplot(aes(x = quality.factor, y = residual.sugar), data = wine) +
  geom_boxplot() + coord_cartesian(ylim = c(0,25))
```

Still, it is not very clear that sugar content impact quality. Since density is
related to alcohol and sugar, it is interesting to plot density against
quality.factor:

```{r echo=FALSE}
ggplot(aes(x = quality.factor, y = density), data = wine) +
  geom_boxplot() + coord_cartesian(ylim = c(min(wine$density), 1.005))
```

It make sense that density is related to quality since alcohol content directly
imapct density.

Some other interesting points are to look at the relationship between pH and 
acidity:

```{r echo=FALSE}
ggplot(aes(x = total.acidity, y = pH), data = wine) +
  geom_point()
```

sulphates and sulfur.dioxide:

```{r echo=FALSE}
ggplot(aes(x = total.sulfur.dioxide, y = sulphates), data = wine) +
  geom_point()
```

sugar and density:

```{r echo=FALSE}
ggplot(aes(x = density, y = residual.sugar), data = wine) +
  geom_point() + 
  coord_cartesian(xlim = c(0.9871,1.005), ylim = c(0,30))
```

alcohol and density:

```{r echo=FALSE}
ggplot(aes(x = density, y = alcohol), data = wine) +
  geom_point() +
  coord_cartesian(xlim = c(0.9871,1.005))
```

# Bivariate Analysis

### Noticeable relationships:

The most notable relationship is between quality and alcohol content. It seems 
that high alcohol content is related to high quality, and since density is 
related to the relative ratio between sugar and alcohol, the higher density the 
lower the alcohol content and therefore the lower the quality.
(It is hard to make high alcohol content wine, since yeasts 
cannot survive in high alcoholic environment.)

I expected pH and total acidity to have a strong correlation since pH is a 
indicator of how acidic the wine is. However, even though there is a correlation
between acidity and pH but not very strong. The reason of this weak correlation 
could be the acid in wine are not strong acid that release all of their protons 
in the solution, which makes sense since wine contains a lot of artaric acid 
from the grapes.

It is also not suprising that sulphates and total sulfur dioxide is not very 
strongly correlated since not all of the sulphates can be transform to sulfur 
dioxide due to the various chemical conditions in different wines.

# Multivariate Plots Section

Now it looks like alcohol content has a moderate positive impact on quality
while other features (acidity, sulfur dioxide, sugar, density and citric acid)
are included into the plot:

First, I classify sulfur dioxide and acidity for better visualization:

```{r, Create_Categories}
wine$total.sulfur.dioxide.bucket = 
  cut(wine$total.sulfur.dioxide, breaks = c(0,100,125,150,175,200,500))
wine$total.acidity.bucket = 
  cut(wine$total.acidity, breaks = c(0,6.5,7,7.5,15))
```

Since I want to use scatter plot to visualize the third variable, here I treat
quality as a numeric variable instead of factor, and I put quality to y-axis:

quality against alcohol and acidity:

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(aes(color = total.acidity.bucket), alpha = 1/3) +
  scale_color_brewer(type = 'div')
```

quality against alcohol and sulfur dioxide:

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(aes(color = total.sulfur.dioxide.bucket), alpha = 1/3) +
  scale_color_brewer(type = 'div')
```

quality against alcohol and sugar:

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), 
       data = subset(wine,
                     residual.sugar < quantile(wine$residual.sugar, 0.99))) +
  geom_jitter(aes(color = residual.sugar), alpha = 1/3) +
  scale_color_distiller(type = 'div', palette = 5)
```

quality against alcohol and density:

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), 
       data = subset(wine, density < quantile(wine$density, 0.99))) +
  geom_jitter(aes(color = density), alpha = 1/3) +
  scale_color_distiller(type = 'div')
```

quality against alcohol and citric acid:

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), 
       data = subset(wine, citric.acid < quantile(wine$citric.acid, 0.99))) +
  geom_jitter(aes(color = citric.acid)) +
  scale_color_distiller(type = 'div', palette = 5)
```

Last, I built a linear model:

```{r}
summary(lm(quality ~ alcohol + total.acidity + total.sulfur.dioxide + 
             residual.sugar + citric.acid, data = wine))
```

# Multivariate Analysis

### Summary:
other than alcohol content, I did not see other noticeable features play an 
important role in quality. The reason could be all the wine samples have above
average quality, and therefore the chemical contents are all in a small range
and even the low quality samples in this dataset have decent chemical property.

The problem of my linear model is that the quality is a discrete variable and
there is not a very clear linear relationship, and other variables does not have
great impact on this model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality.factor, y = alcohol), data = wine) +
  geom_boxplot() +
  xlab('Quality') + ylab('Alcohol (% by volume)') +
  ggtitle('Boxplot: Alcohol vs Quality')
```

### Description One

This plot shows the relationship between quality and alcohol content where the
overall trend is that high quality wines seem to have a relatively high the
alcohol content.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = density, y = alcohol), data = wine) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = 'lm') +
  xlab('Density (g/cm^3)') + ylab('Alcohol (% by volume)') +
  coord_cartesian(xlim = c(0.9871,1.005), ylim = c(7.5,15)) + 
  ggtitle('Scatterplot: Alcohol vs Density')
```

### Description Two

This plot shows the strong correlation between alcohol content and density which
interprets the weak correlation between density and quality. Alcohol content is
tastable while density is not, and therefore alcohol content is more likely to
be the real reason that impacts quality.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(aes(color = total.acidity.bucket), alpha = 1/3) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Total Acidity',
                                          override.aes = list(alpha = 1))) +
  ylab('Quality') + xlab('Alcohol (% by volume)') +
  ggtitle('Scatterplot: Quality vs Alcohol and Total acidity')
```

### Description Three

It is surprising that acidity does not have a noticeable impact on quality. Even
some of the high quality wines have high acidity.

------

# Reflection

The dataset is very tidy which makes it very easy to do analysis. All of the
variables are close to normal distribution which means I do not need to do
variable transformation for normal distribution.

The struggle is that most of the variables do not have even moderate correlation
with the feature that I am interested in (quality). It seems like it is hard to
just use these chemical properties to predict wine quality. There might be other
chemicals that are more related to wine taste or even some other features of
wine have impact on taste.

In the future, this dataset could include more chemical properties and have more
precise quality classification. More high quality wines (above 7) and low
quality wines (below 5) could also help better distinguish wine quality and
build models.

